---
title: "ds202_lab4"
author: "Lisa McDaniel"
date: "10/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<font size="5"> Loading Data:

```{r message=FALSE}
library(readxl)
library(tidyverse)

defense <- read_excel('C:/Users/lisa.mcdaniel/DS202/ds202lab4/cyclonesFootball2019.xlsx',
                      sheet="Defensive")

offense <- read_excel('C:/Users/lisa.mcdaniel/DS202/ds202lab4/cyclonesFootball2019.xlsx',
                      sheet="Offensive")

bio <- read_excel('C:/Users/lisa.mcdaniel/DS202/ds202lab4/cyclonesFootball2019.xlsx',
                      sheet="Biography")
```

<font size="5"> Part One: Cleaning Data

<font size="3"> 1.	The player names and opponent names are factors.
    
```{r message=FALSE}

def1 <- defense %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent))

off1 <- offense %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent))

bio1 <-  bio %>%
  mutate(Name = as.factor(Name))

```

<font size = "3"> 2.	All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)

```{r message=FALSE}

defClean <- def1 %>%
  mutate(across(.cols = (Tackles_Solo:Pass_PB), .fns = (as.numeric)))

offClean <- off1 %>%
  separate(`Passing_CMP-ATT`, c('Passing_CMP', 'Passing_ATT')) %>%
  mutate(across(.cols = (Rushing_ATT:Passing_INT), .fns = (as.numeric)))

bio2 <- bio1 %>%
  mutate(Weight = as.numeric(Weight))

```

<font size = "3"> 3.	Change the Height column in biogrpahy into numeric. (Hint: Use a unit that works well).

Converted the height to inches; column heading *Height_in*.  Only showing the first 6 rows of the *bioClean* dataframe.

```{r message=FALSE}
bioClean <- bio2 %>%
 separate(Height, c('feet', 'inches'), sep = '-') %>%
 mutate(feet = as.numeric(feet), inches = as.numeric(inches)) %>%
 mutate(Height_in = (12*feet + inches)) %>%
 select(-feet, -inches)

head(bioClean)
```

<font size = "3"> Cleaned Dataframes

```{r meassage=FALSE}
str(defClean)
str(offClean)
str(bioClean)
```

<font size="5"> Part Two: Tidying

<font size = 3">  1.	Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.

Saved the new dataframe as *defClean1*; only showing the first 6 rows of the *defClean1* dataframe in the output.

```{r message=FALSE}

defClean1 <- defClean %>%
  pivot_longer(`Tackles_Solo`:`Pass_PB`, names_to='Def_Skill', values_to='Stat')

head(defClean1)
```

<font size = 3">  2.	Compare the distributions of the defensive statistics. What defensive skills are rare?

Created a barplot showing the defensive skills and arranged the information in ascending order.  The 3 rarest defensive skills are interception, fumble recovery and fumble, respectively, all are related to turnovers.

```{r message=FALSE}

skill <- c("Turnover_INT", "Turnover_FR", "Turnover_FF", "Tackles_Sack", 
           "Pass_QBH", "Pass_PB", "Tackles_TFL", "Tackles_ASST", "Tackles_Solo")

ggplot(defClean1, aes(x=Def_Skill, y=Stat, fill=Def_Skill)) + 
  geom_bar(stat= 'identity') + scale_x_discrete(limits = skill) + 
  theme(axis.text.x = element_text(angle = 90)) + ylab("Count") + 
  xlab("Defensive Skill") + theme(legend.position="none")
```

<font size = 3">  3.	Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.

From the scatterplot it appears that ISU had a better defense against Notre Dame.  There doesn't appear to be a identifiable trend as to why there was an increase in tackles other then if the player participated in the Iowa game, their number of solo tackles increased during the Notre Dame game.

```{r message=FALSE}

defClean2 <- defClean1 %>%
  filter(Opponent_Opponent == c("Iowa", "Notre Dame"), 
         Def_Skill == "Tackles_Solo")

ggplot(data=defClean2, aes(y=Stat, x=Name, color=Opponent_Opponent)) + 
  geom_point() + theme(axis.text.x = element_text(angle = 90)) + 
  ylab("Solo Tackles") + xlab('Tackler') + labs(color="Opponent")
```

<font size = 3">  4.	Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r message=FALSE}
bioClean1 <- bioClean %>%
  separate(Hometown, c('City', 'State'), sep = ',')

head(bioClean1[, c("City", "State")], n=5)
```

5.	How many players are there from each state?

I created a dataframe called *St1* to store the state and number of players from each state, by state in alphabetical order.  I'm only showing the first 6 rows in the output.

```{r message=FALSE}
St1 <- bioClean1 %>%
  select(Name, State) %>%
  count(State)

head(St1)  
```

<font size="5"> Part Three: Joining Data Frames

<font size = 3">  1.	Find from which states do the ISU players come. Create an appropriate summary.

I used *group_nest()* to get individual tables for each state containing the name of the players from that state.  The individual tables are stored in *St2*.  The output just shows the first 5 tables stored in *St2*.  The number of rows, in the first 5 tables in *St2*, corresponds to the number of players stored in *St1* from question 5 for each state.  Just for fun I also created a bar plot showing the number of players from each state.  Not surprisingly, most of the players are from Iowa.  

```{r message=FALSE}
St2 <- bioClean1 %>%
  select(Name, State) %>%
  group_nest(State) %>%
  as_tibble()
 
head(St2, n=5)

ggplot(data=St1, aes(x=State,  y=n, fill=State)) + 
  geom_bar(stat= 'identity') +  ggtitle("Number of Players Per State") + 
  theme(plot.title = element_text(hjust = 0.5)) +  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90)) + ylab("Number of Players") + 
  xlab("State")
```

2.	Is there a relationship between the player statistics and their positions?

Defense:  Using facet wrap, I created individual bar plots, by defensive position, to show there is a relationship between player statistics and position.  For example, you can tell the Defensive Back and Linebacker both dominate tackle related defense skills.
```{r message=FALSE, fig.width=20, fig.height=15}
bioClean3 <- bioClean %>%
  select(Name, Position)

defPos <- defClean %>%
  left_join(bioClean3) %>%
  select(-Opponent_Opponent, -Name) %>%
  group_by(Position) %>%
  filter(!is.na(Position)) %>%
  summarise_at(vars(Tackles_Solo:Pass_PB), sum, na.rm=TRUE) %>%
  pivot_longer('Tackles_Solo':'Pass_PB', names_to='Def_Skill', 
               values_to='Stat')
  
ggplot(defPos, aes(Position, Stat, fill=Def_Skill)) + 
  geom_bar(stat = "identity", position = 'dodge', width = .75) +  
  ggtitle("Defensive Skill by Position") + theme(plot.title = element_text(size = 30)) +
  theme(plot.title = element_text(hjust = 0.5))  + theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90)) + ylab("") + 
  theme(axis.text = element_text(size = 20)) + theme(axis.title = element_text(size = 30)) + 
  xlab("Defensive Skill") + facet_wrap(~Def_Skill) + theme(strip.text.x = element_text(size = 20))
```

Offense: Using facet wrap, I created individual bar plots, by offensive position, to show there is a relationship between player statistics and position.  For example, you can tell the quarterback dominates passing related offense skills.
```{r message=FALSE, fig.width=20, fig.height=15, warning=FALSE}
offPos <- offClean %>%
  left_join(bioClean3) %>%
  select(-Opponent_Opponent, -Name) %>%
  group_by(Position) %>%
  filter(!is.na(Position)) %>%
  summarise_at(vars(Rushing_ATT:Passing_INT), mean, na.rm=TRUE) %>%
  pivot_longer('Rushing_ATT':'Passing_INT', names_to='Off_Skill', 
               values_to='Stat')

ggplot(offPos, aes(Position, Stat, fill=Off_Skill)) + 
  geom_bar(stat = "identity", position = 'dodge', width = .75) +  
  ggtitle("Offensive Skill by Position") + theme(plot.title = element_text(size = 30)) +
  theme(plot.title = element_text(hjust = 0.5))  + theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90)) + ylab("") + 
  theme(axis.text = element_text(size = 20)) + theme(axis.title = element_text(size = 30)) +
  xlab("Offensive Skill") + facet_wrap(~Off_Skill) +theme(strip.text.x = element_text(size = 20))
```

3.	How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

After filtering for *Brock Purdy* and removing several columns that weren't relevant to the quarterback perfromance, I used the information from https://www.dummies.com/sports/football/offense/how-to-read-a-quarterbacks-statistics/ to calculate an overall quarterback *Rating* which I arranged in ascending order.  Next I added a column that recorded whether the game was a win or loss for Iowa State and trimmed the dataframe to only contain *Name*, *Opponent_Opponent*, *W_L* and *Rating* columns.  It appears that if Bock Purdy's *Rating* was less then 100, there was a good chance that Iowa State would lose; if the *Rating* was over 100 there was a good chance that Iowa State would win.  Of course, there are always an outlier or two. 

```{r message=FALSE}
offClean2 <- offClean %>%
  filter(Name == "Purdy, Brock") %>%
  select(-Receiving_REC, -Receiving_TD, -Receiving_YDS) %>%
  mutate(Rating = (((((Passing_CMP/Passing_ATT)-0.3)/0.2) + 
           (((Passing_YDS/Passing_ATT)-3)/4) + ((Passing_TD/Passing_ATT)/0.05) + 
  ((0.095-(Passing_INT/Passing_ATT))/0.04))*100)/6) %>%
  arrange(Rating) %>%
  mutate(W_L = ifelse(Opponent_Opponent %in% c('UNI', 'ULM', 'TCU', 'West Virginia', 'Texas Tech', 'Texas', 'Kansas'), 'W', 'L')) %>%
  select(Name, Opponent_Opponent, W_L, Rating)
 
offClean2
```

4.	Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.

Defense:  There were 28 players that played during both the 2018 and 2019 seasons,  O'Rien Vance and Zach Petersen appear to show the most improvement in several defensive skills; JaQuan Baiey appears to be getting worse.
```{r message=FALSE, fig.width=20, fig.height=15}

defense2 <- read_excel('C:/Users/lisa.mcdaniel/DS202/ds202lab4/cyclonesFootball2018.xlsx',
                      sheet="Defensive")

offense2 <- read_excel('C:/Users/lisa.mcdaniel/DS202/ds202lab4/cyclonesFootball2018.xlsx',
                      sheet="Offensive")

defClean2N <- defense2 %>%
  select(Name) %>%
  distinct()

defClean3N <- defClean %>%
  select(Name) %>%
  distinct()

DefNames <- inner_join(defClean2N, defClean3N)

defClean2 <- defense2 %>%
  select(-Opponent_Opponent) %>%
  mutate(Name = as.factor(Name)) %>%
  mutate(across(.cols = (Tackles_Solo:Pass_PB), .fns = (as.numeric))) %>%
  group_by(Name) %>%
  summarise_at(vars(Tackles_Solo:Pass_PB), sum, na.rm=TRUE) %>%
  mutate(Year = '2018') %>%
  right_join(DefNames) %>%
  pivot_longer(`Tackles_Solo`:`Pass_PB`, names_to='Def_Skill', 
               values_to='Stat')

defClean3 <- defClean %>%
  select(-Opponent_Opponent) %>%
  group_by(Name) %>%
  summarise_at(vars(Tackles_Solo:Pass_PB), sum, na.rm=TRUE) %>%
  mutate(Year='2019') %>%
  right_join(DefNames) %>%
  pivot_longer(`Tackles_Solo`:`Pass_PB`, names_to='Def_Skill', 
               values_to='Stat')

defComp <- full_join(defClean2, defClean3)
  
head(defComp)

ggplot(defComp, aes(Def_Skill, Stat, fill=Year)) + 
  geom_bar(stat = "identity", position = 'dodge', width = 0.5) +  
  ggtitle("ISU Defense Progress Between 2018 & 2019") + 
  theme(plot.title = element_text(size = 30)) + 
  theme(plot.title = element_text(hjust = 0.5)) +  theme(legend.position="none") +
  theme(axis.text = element_text(size = 20)) + theme(axis.title = element_text(size = 30)) +
  theme(axis.text.x = element_text(angle = 90)) + ylab("") + 
  xlab("Defensive Skill") +  facet_wrap(~Name, 8) + theme(strip.text.x = element_text(size = 20))

```


Offense:  There were 8 players that played during both the 2018 and 2019 seasons.  Brock Purday appears to show the most improvement.
```{r message=FALSE, fig.width=20, fig.height=15}
offClean2N <- offense2 %>%
  select(Name) %>%
  distinct()

offClean3N <- offClean %>%
  select(Name) %>%
  distinct()

OffNames <- inner_join(offClean2N, offClean3N)

offClean2 <- offense2 %>%
  select(-Opponent_Opponent) %>%
  mutate(Name = as.factor(Name)) %>%
  separate(`Passing_CMP-ATT`, c('Passing_CMP', 'Passing_ATT')) %>%
  mutate(across(.cols = (Receiving_REC:Passing_INT), .fns = (as.numeric))) %>%
  group_by(Name) %>%
  summarise_at(vars(Receiving_REC:Passing_INT), sum, na.rm=TRUE) %>%
  mutate(Year = "2018") %>% 
  right_join(OffNames) %>%
  pivot_longer('Receiving_REC':'Passing_INT', names_to='Off_Skill', 
               values_to='Stat')

offClean3 <- offClean %>%
  select(-Opponent_Opponent) %>%
  group_by(Name) %>%
  summarise_at(vars(Receiving_REC:Passing_INT), sum, na.rm=TRUE) %>%
  mutate(Year='2019') %>%
  right_join(OffNames) %>%
  pivot_longer('Receiving_REC':'Passing_INT', names_to='Off_Skill', 
               values_to='Stat')

offComp <- full_join(offClean2, offClean3)

ggplot(offComp, aes(Off_Skill, Stat, fill=Year)) + 
  geom_bar(stat = "identity", position = 'dodge', width = 0.5) +  
  ggtitle("ISU Offense Progress Between 2018 & 2019") + 
  theme(plot.title = element_text(size = 30)) +
  theme(plot.title = element_text(hjust = 0.5)) +  theme(legend.position="none") +
  theme(axis.text = element_text(size = 20)) + theme(axis.title = element_text(size = 30)) +
  theme(axis.text.x = element_text(angle = 90)) + ylab("") + 
  xlab("Defensive Skill") +  facet_wrap(~Name) +theme(strip.text.x = element_text(size = 20))
```
